Research Experience

### 2010-04-28 doc/thesis/thesis.tex
emph{sort} of done with the lit review, but i need to review it...

\subsection{Previous Studies of Parameters}
There have been quite a few efforts to quantify the parameters of the
various BOLD models.  Although \cite{Buxton1998} and \cite{Friston2000}
both proposed physiologically reasonable values for the model parameters, 
\cite{Friston2002} was the first paper to calculate the parameters based 
on actual FMRI data. In that paper, Friston et. al. used a variation of
Expectation Maximization to find a normal distribution for the parameters:
\begin{eqnarray}
\epsilon &=& N(.54 , .1 ^2 )  \nonumber \\
\tau_s & =&  N(1.54, .25^2)   \nonumber \\
\tau_f & =&  N(2.46, .25^2)   \nonumber \\
\tau_0 & =&  N(.98 , .25^2 )   \nonumber \\
\alpha & =&  N(.33 , .45^2 )   \nonumber \\
E_0   & =&  N(.34 ,  .1 ^2 )   \nonumber \\
V_0  & = &  .03 (not\ estimated) \nonumber
\end{eqnarray}
Since then, several other methods of have been used to calculate
BOLD parameters from FMRI timecourses. In \cite{Riera2004}, a maximum
likelihood method for innovation processes was used as described by
\cite{Ozaki1994}
later \cite{Hu2009} used an Unscented Kalman Filter to confirm these findings.
Both \cite{Hu2009} and \cite{Friston2002} came to results very similar to the
expected values.
On the other hand \cite{Johnston2008} used a hybrid particle filter Expectation-
Maximization algorithm to calculate parameters and state simultaneously.
His derived parameter distributions differed greatly from Friston et. al.:
\begin{eqnarray}
epsilon &=& .069 \pm .014    \nonumber \\
tau_s & =& 4.98 \pm 1.07  \nonumber \\
tau_f & =& 8.31 \pm 1.51   \nonumber \\
tau_0 & =& 8.38 \pm 1.5    \nonumber \\
alpha & =& .189 \pm .004   \nonumber \\
E_0   & =& .635 \pm .072     \nonumber \\
V_0   & =& .0149 \pm .006     \nonumber 
\end{eqnarray}
%here
On the one hand, Johnston et. al. is a pseudo-maximum likelihood estimator, and 
it depends very little on the priors. On the other hand, the results are very 
different from \cite{Buxton1998}, \cite{Hu2009}, \cite{Friston2000}, and 
\cite{Friston2002}. All four of the papers just mentioned all have a common root,
and all used essentially the same priors, and all use Bayesian estimation
that depends greatly on the priors. In this case, it is possible that priors
biased the results, preventing accurate measurements of the parameters. Its also
possible that the parameters don't lend themselves to learning very well. 
It would be interesting to see what
effect different priors would have on the results in \cite{Friston2002}.

In the end, its possible that the balloon model may be as complex as the
measurement rate and error size will allow before the error becomes dominated 
by model variance, so proving that
Behzadi's Arterial Compliance model is quantitatively better than Buxton's 
advanced balloon model \cite{Buxton2004} or that either is better than 
the extremely detailed model found in \cite{Zheng2005} may be impossible. 
The reality is that
measurement noise alone may dominate any difference between these models.

\subsection{Noise}
Thus, despite some discrepencies, the cause of the BOLD signal is
relatively well known.  but FMRI doesn't detect this happening in one neuron,
but rather as the 
aggregate over millions of cells. Though local neurons act
"together" (i.e. around the same time), the density of neurons, the
density of capillaries, and slight differences in activation across 
a particular voxel can all lead to signal attenuation or noise. 
A particularly insidious type of noise present in FMRI is a low frequncey
drift, first characterized by a Weiner process in \cite{Riera2004}. 
Though not present in all regions, it is prevalent enough to cause significant
inference problems \cite{Tanabe2002}. It is still not
clear where exactly this noise comes although it is possible it is 
the result of magnets heating up, or some distortion in magnetic
fields \cite{Smith2007}. It is clear that this drift signal is not soley
due to a true physiological
effects however, given its presence in cadavers and phantoms \cite{Smith1999}.
Additionally, since it is standard operating procedure to perform
coregistration between volumes, it is unlikely that movement was
the cause of the drift either.

\section{Current Techniques}
\subsection{Basic Statistical Parametric Mapping}
The most basic method of analyzing FMRI data is through a basic T-test
between "resting state" and "active state" samples. This is done by 
taking the average and variance of the inactive period, and the 
active period separately then treating them both as gaussian distributions.
If they are in fact Gaussian distributions, then a basic t-test will
give the likelihood that the samples came from the same distribution
(the p-value). Of course, this test is fraught with problems; even if
the drift mentioned earlier has been removed, there is little reason
to believe that the noise is Gaussian, or even stable. Additionally, 
even if the noise were Gaussian, a t-test with a p-value of .05 over
5000 or more samples is on average going to generate $.05*5000$ false
positives. To compensate for this, bonferoni correction, also known as
multiple comparison tests are performed; essentially p-values are 
divided by the number of independent
tests being run. This, however, leads to extremely low p-values, so
low that it would be impossible for any biological system to satisfy. To
compensate, a Gaussian kernel is applied to the image, thus reducing
variance (and thus separating the active and inactive distributions)
as well as decreasing the effective number of voxels. Since t-tests are
now no longer being applied to n <I need to define n> independent voxels,
the factor by which the p-value must be divided by can be decreased.
<Do I need to mathematically define all this?> The derivation and application
of random field theory, and its use can be found in various papers \cite{univ_mult_rft}.

\subsection{General Linear Model}
The most used form of FMRI analysis is still based on Statistical Parametric
Mapping, but is able to account for several different levels or types
of stimulus. By adding a General Linear Model to the analysis, the
output signal timeseries (what FMRI detects) is regressed over the weighted
sum of the various confound's timeseries. The equation for a general linear
model is then
\begin{equation}
Y(t) = X(t)\beta + \epsilon(t)
\end{equation}
where $Y(t)$ is the smoothed or detrended timeseries of measurements,
$X(t)$ is a row vector of stimuli, $\beta$ is a column vector of weights,
and $\epsilon$ is the error. Thus for every time, the measurement is
assumed to be a weighted sum of the inputs plus some error. The calculation
of $\beta$ then is merely a gradient descent search to minimize the
mean squared error.

@@

Of course, a square wave input is not going to result in a square wave
in the activation of brain regions. Thus, various methods are used to 
smooth $X(t)$ through time, and bandlimit the input. The best technique
is convolving the stimulus input with a hemodynamic response function,
which mimicks the basic shape of BOLD activation, including a delay
due to rise time and fall time. This hemodynamic signal is static however,
so every region of the brain gets the same design matrices (X(t)), 
although the weights of various stimulus or confounds are allowed to vary.

@@

Ultimately, activation due to a particular stimuli is decided by the 
$\beta$ value corresponding to that stimuli's column of $X(t)$.
<Need to check this>
The null hypothesis as to whether the outcome was random is then
based on a t-test of the $\epsilon(t)$ timeseries.

\subsection{Whats wrong with these techniques}
There are a few problems with the techniques mentioned in the 
previous sections. First, they essentially ignore prior knowledge about
the system. Although the most advanced form of the general linear model includes
a "Hemodynamic Response Function," that hemodynamic response function is
static across every region of the brain. It is well known that capillary beds
are not uniform and so blood perfusion cannot possible be static across the
brain. Thus, if extra information were available a-priori, that information
could not be incorporated without modifications to the General Linear Model.
Similarly heart rate could not be added either. It would obviously be advantageous
to have true physiological parameters as entry points for these various other
model parameters. The physiological models for the BOLD signal are quite good
and based on realistic physics. While the exact connection between a stimulus
and the flow inducing signal is not precisely known, model fits are actually
quite good \cite{nonlinearmodels}. Regardless, being based on some real 
physiological 
parameter would allow for the establishment of reasonable priors and decrease
model variance without breaking a sweat. Of course, using real parameters 
has the additional
bonus of potentially providing information about physical pathologies. It
is quite possible that physical properties such as decreased compliance of
blood vessels could indicate a neurological condition that is not easily
seen in a T1 or T2 map. In essence, this could make FMRI a much more 
useful clinical tool than it is now. The other problem with linear models
is that they are a linear fit to a nonlinear signal. It is not uncommon
for data to be thrown out in FMRI studies because no significant activation
has been seen. However, if, for whatever reason, the BOLD response
was acting more nonlinear than in other patients it would be completely
possible for SPM to miss that activation.

@@

Secondly, these methods are still based on t-tests, which notoriously
lack robustness to non-Gaussian noise. While different techniques 
exist for imposing Gaussianity, those techniques are incapable of
discriminating noise from signal. There is no way to know how much 
signal is removed by various smoothing techniques, or even if entire 
regions have been smoothed into oblivion. Instead of extensively filtering data to 
remove noise, the analysis method itself must be robust a wide range of noise,
which is why we propose here the use of particle filters.

>>>>

\subsection{Previous Studies of Parameters}
There have been quite a few efforts to quantify the parameters of the
various BOLD models.  Although \cite{Buxton1998} and \cite{Friston2000}
both proposed physiologically reasonable values for the model parameters, 
\cite{Friston2002} was the first paper to calculate the parameters based 
on actual FMRI data. In that paper, Friston et. al. used a variation of
Expectation Maximization to find a normal distribution for the parameters:
\begin{eqnarray}
\epsilon &=& N(.54 , .1 ^2 )  \nonumber \\
\tau_s & =&  N(1.54, .25^2)   \nonumber \\
\tau_f & =&  N(2.46, .25^2)   \nonumber \\
\tau_0 & =&  N(.98 , .25^2 )   \nonumber \\
\alpha & =&  N(.33 , .45^2 )   \nonumber \\
E_0   & =&  N(.34 ,  .1 ^2 )   \nonumber \\
V_0  & = &  .03 (not\ estimated) \nonumber
\end{eqnarray}
Since then, several other methods of have been used to calculate
BOLD parameters from FMRI timecourses. In \cite{Riera2004}, a maximum
likelihood method for innovation processes was used, as described by
\cite{Ozaki1994}. \cite{Ozaki1994} uses a similar construction to a 
Kalman filter, to break the time series into a series of innovations,
for which Maximum Likelihood was performed. While this is in some since the
"right" way to find the solution, it comes with several caveats. First, every
step in parameter space requires a recalculation of all the state variables. With
two or three parameters this is fine, more than that, and calculations could go on
indefinitely. Second, it still assumes the parameters and noise are Gaussian, and
will only be optimal in that case. Third, depending on the nonlinearities present
in the system, local minima may be extremely common. Later \cite{Hu2009} used an 
Unscented Kalman Filter over all the parameters and state variables to find the 
parameter set/variable time series. While this method has the drawback of not necessarily
being optimal, unfortunately there is no general optimal solution to non-linear non
Gaussian models. Hu et. al.'s technique also will run significantly faster than
ML based techniques, since it does not require recalculating the entire timeseries
for every step in parameters. Both \cite{Hu2009} and \cite{Friston2002} came to results 
very similar to the expected values stated in \cite{Buxton1998} and \cite{Friston2000}.
One potential problem with all these techniques are that the depend heavily on the
priors. The starting point of the parameters could have a huge impact on the
results, and while one can be hopeful that this isn't the reason for the agreement
between \cite{Friston2000} and later results, there is no way to know.

In \cite{Jonston2008}, a hybrid particle filter/gradient
descent algorithm was used to simultaneously derive the static (classicly called
parameters) and dynamic parameters (classically known as state variables).
Essentially a particle filter is used to calculate the state variables at some
time, and then the estimated distribution of the particles was used to find
the most likely set of parameters that would give that distribution of state variables.
\cite{Johnston2008} comes to a very different set of parameter estimates as compared
to the original \cite{Friston2000} guesses:
\begin{eqnarray}
epsilon &=& .069 \pm .014    \nonumber \\
tau_s & =& 4.98 \pm 1.07  \nonumber \\
tau_f & =& 8.31 \pm 1.51   \nonumber \\
tau_0 & =& 8.38 \pm 1.5    \nonumber \\
alpha & =& .189 \pm .004   \nonumber \\
E_0   & =& .635 \pm .072     \nonumber \\
V_0   & =& .0149 \pm .006     \nonumber 
\end{eqnarray}
Notably, the time constants are significantly longer. This could be a result of
some preprocessing to the stimulus timeseries performed in \cite{Friston2002} and
later works but not in \cite{Johnston2008}, or it could be that \cite{Johnston2008}
depended less on the priors.

Its possible, although unlikely that the balloon model is not possible to learn
without detrimental cost in variance error. I do not think this is the case, however,
and I will work to dispell this possibility in the results with simulated time
series.

\subsection{Noise}
Thus, despite some discrepencies, the cause of the BOLD signal is
relatively well known.  but FMRI doesn't detect this happening in one neuron,
but rather as the 
aggregate over millions of cells. Though local neurons act
"together" (i.e. around the same time), the density of neurons, the
density of capillaries, and slight differences in activation across 
a particular voxel can all lead to signal attenuation or noise. 
A particularly insidious type of noise present in FMRI is a low frequncey
drift, first characterized by a Weiner process in \cite{Riera2004}. 
Though not present in all regions, it is prevalent enough to cause significant
inference problems \cite{Tanabe2002}. It is still not
clear where exactly this noise comes from, although it is possible it is 
the result of magnets heating up, or some distortion in magnetic
fields \cite{Smith2007}. It is clear that this drift signal is not soley
due to a physiological
effects, given its presence in cadavers and phantoms \cite{Smith1999}.
Additionally, since it is standard operating procedure to perform
coregistration between volumes, it is unlikely that movement was
the cause of the drift either (otherwise the effect would not be 
widely reported).

\section{Current Techniques}
\subsection{Basic Statistical Parametric Mapping}
Although not strictly the same thing as parameter calculation from 
FMRI, activation detection is very similar. In fact, estimation of 
parameters is somewhat a generalization of the idea of activation detection.
Thus, it is important to draw a comparison between the methods proposed
in this thesis with existing methods of activation detection.

The most basic method of analyzing FMRI data is through a standard T-test
between "resting state" and "active state" samples. This is done by 
taking the average and variance of the inactive period, and the 
period during which the stimulus was activate separately then treating 
them both as gaussian distributions.
If they are in fact Gaussian distributions, then a basic t-test will
give the probability that the samples came from the same distribution
(the null hypothesis). Of course, this test is fraught with problems; even if
the drift mentioned earlier has been removed, there is little reason
to believe that the noise is Gaussian, or even stable. Additionally, 
even if the noise were Gaussian, a t-test with a p-value of .05 over
50000 or more samples is on average going to generate $.05*50000$ false
positives. To compensate for this, bonferoni correction, also known as
multiple comparison tests are performed; essentially p-values are 
divided by the number of independent
tests being run. This, however, leads to extremely low p-values, so
low that it would be impossible for any biological system to satisfy. To
compensate, a Gaussian kernel is applied to the image, thus reducing
variance (and thus separating the active and inactive distributions)
as well as decreasing the effective number of voxels. Since t-tests are
now no longer being applied to n <I need to define n> independent voxels,
the factor by which the p-value must be divided by can be decreased.
<Do I need to mathematically define all this?> The derivation and application
of random field theory, and its use can be found in various papers \cite{univ_mult_rft}.

\subsection{General Linear Model}
The most used form of FMRI analysis is Statistical Parametric
Mapping, but is able to account for several different levels or types
of stimulus (see \cite{Hofmann1997}). By using hierarchical models
the output signal timeseries is considered the weighted
sum of the various input timeseries. Essentially every experimental factor
is considered as another level inputs (ex. which patient is an input). 
The equation for a general linear model is then
\begin{equation}
Y(t) = X(t)\beta + \epsilon(t)
\end{equation}
where $Y(t)$ is the smoothed or detrended timeseries of measurements,
$X(t)$ is a row vector of input, $\beta$ is a column vector of weights,
and $\epsilon$ is the error. Thus for every time, the measurement is
assumed to be a weighted sum of the inputs plus some error. The calculation
of $\beta$ then performed using a maximum likelihood or gradient descent search 
to minimize the error.

@@

It is well known of course that a square wave stimulus does not result in a square wave
in the activation of brain regions. Thus, various methods are used to 
smooth $X(t)$'s related to actual input (as opposed to experimental confounds
such as patient number, age etc) through time, and bandlimit the input. 
The best technique is convolving the stimulus input with a hemodynamic 
response function, which mimicks the basic shape of BOLD activation, including a delay
due to rise time and fall time. The downside of this method, however is that 
the hemodynamic signal is static, meaning the same hemodynamic function is
used for every region of the brain, and $Y(t)$ must be a constant
scalar multiple of $X(t)$ to be appropiately identified.

@@

The GLM is extremely powerful at determining the linear dependence of
a set of regressors on the output. Unfortunately, there is significant evidence
that many of these dependencies are nonlinear, which means they may be
difficult or impossible to detect using linear techniques. Additionally, 
the lack of flexibility in delay variables certainly hinders SPM's
ability to locate voxels that do not conform to the canonical hemodynamic 
response function. The very fact that different hemodynamic response functions
work better for different regions/experiments is a testament to the need
for increased flexibility.

Linear models also ignore prior knowledge about the system. Combined FMRI
CBF or CBV imaging methods are available and could shed much light on
neural activation; however there is no way of combining that data into a
linear model. Although the most advanced form of the general linear model includes
a Hemodynamic Response Function, it is static across every region of the brain;
yet it is well known that capillary beds are not uniform and so blood perfusion 
cannot possibly be static across the brain. Its possible that activation is
every bit as strong in regions that aren't found by SPM, but that there are
nonlinearities that do not match the canonical HRF; resulting in signficant bias error. 
It is not uncommon
for data to be thrown out in FMRI studies because no significant activation
has been seen. This practice highlights the fact that strictly linear
approaches are insufficiently flexible to account for relatively common
variations of neural activation.

@@

In general activation detection type methods also don't have the ability 
to find pathologies based on state variables or parameters.  It
is quite possible that physical properties such as decreased compliance of
blood vessels could indicate a neurological condition that is not easily
seen in a T1 or T2 map. In essence, this could make FMRI a much more 
useful clinical tool than it is now.

Finally, all these techniques require the noise to be Gaussian to reach
an optimal solution. In fact there is no known optimal solution to
nonlinear models with non-Guassian noise, so obviously some assumptions
are going to have to be made to reach a solution. However, it would be
nice to have an algorithm that is robust to these effects, and could
still give a good solution when Gaussianity is violated.
